---
title: "sim.5.28"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(mvtnorm)
```

## One outcome/ One predictor

Assume we have $p$ covariates, $(X_{jt}$ $j=1,\ldots,p$, $t=0,1)$ , measured before $A_0$ ($t=0$) and also before $A_1$ ($t=1$). Suppose that treatment decisions depend on all covariates and direct effects of treatment on each covariate are unknown. Each element of $X_{jt}$ is sampled from a normal distribution.

\begin{align}
\sigma^2 &\sim \text{IG}(1/2,1/2) \\
P(\theta) &= {1}(0<\theta<1) \\
p(\gamma_i=1 \vert \theta) &= \theta \\
p(\beta_i \vert \gamma_i) &= \gamma_i \mathcal{ N}(0,\nu_1) + (1-\gamma_i)\mathcal{N}(0,\nu_0) \\
p(\mathbf{x_{j1}} \vert \boldsymbol{\beta}, \mathbf{X}) &= \mathcal{N}(\mathbf{X} \beta, \sigma^2 \mathbf{I}) \\
\end{align}

The design matrix $\mathbf{X}$ consists of the previous measurement of $\mathbf{x}_{i1}$ as well as the previous treatment, $\mathbf{A}_0$.

$$p(\theta, \gamma, \beta, \sigma^2 \vert \textbf{Data})$$

\begin{align}
p(\theta \vert \gamma) &\sim \text{Beta}(\sum_i \gamma_i + 1, p - \sum_i \gamma_i + 1) \\
p(\gamma_i \vert \beta, \theta) &= \frac{\theta p(\beta \vert \gamma=1)}{\theta p(\beta \vert \gamma=1)+(1-\theta) p(\beta \vert \gamma=0)} \\
&= \frac{\theta N(0,\nu_1)}{\theta N(0,\nu_1) + (1-\theta) N(0, \nu_0)} \\
p(\beta \vert \gamma) &\sim N( (\sigma^{-2}) \mathbf{M}^{-1} X^\top y, \mathbf{M}^{-1}) \\
p(\sigma^2 \vert \cdot) &\sim \text{IG}((n+1)/2, (\text{ss}+1)/2)
\end{align}

where $$\mathbf{M} =(\sigma^{-2}X^\top X + D(\gamma)^{-1})$$


```{r simulate data}
### simulate data 
### 2 covariates, A_0 has a direct effect on one but not the other
### y depends on the last measured treatment and X1

set.seed(3)
n = 1000
u = rnorm(n, mean = 0, sd = 2)
Z1 = rnorm(n)
X1 = rnorm(n)

A1 = rbinom(n, 1, prob = plogis(Z1 + X1)) # higher probability of treatment when X's are >

Z2 = rnorm(n, mean = 0.8 * Z1 + u)
X2 = rnorm(n, mean = 0.7 * X1 - A1 + u) # responds to treatment

A2 = rbinom(n, 1, prob = plogis(Z2 + X2)) # higher probability of treatment when X's are >

Y = rnorm(n, mean = 0.7 * X2 - A2 + u)

dat = data.frame(Z1, Z2, X1, X2, A1, A2, Y)
```

```{r gibbs sampling fn}
# Gibbs sampler for new parameter values
# input: old parameter values (theta, gamma, beta, sigsq), data (design, response)
# output: list of new sampled values for each parameter

sample.new = function(theta.old, gamma.old, beta.old, sigsq.old, design, resp) {
  
  # sample new theta
  # assumpion: U(0,1) prior on theta/ binomial distribution on gamma (beta - binomial)
  theta.new = rbeta(1, shape1 = sum(gamma.old) + 1  ,shape2 = p - sum(gamma.old) + 1)
  
  # sample new gamma 
  # assumption: independent elements, fixed nu_0, nu_1
  p = ncol(design)
  pg = rep(0,p) # hold the gamma probabilities
  
  for(pred in 1:p) {
    a = (theta.new * dnorm(beta.old[pred],mean = 0, sd = nu_1))
    b = (1-theta.new) * dnorm(beta.old[pred],mean = 0, sd = nu_0)
    pg[pred] = a/(a+b)
  }
  
  gamma.new = rbinom(p, 1, prob = pg) # now, sample based on posterior probabilities
  
  # sample new beta
  # assume scale mixture prior and normal likelihood for response with sigma = sigma^2 * I
  M = (t(design) %*% design)*sigsq.old^-1 + diag(ifelse(gamma.new == 1, 1/nu_1, 1/nu_0)) 
  Minv = solve(M)
  meanvec = sigsq.old^-1 * Minv %*% t(design) %*% resp
  beta.new = rmvnorm(1, mean = meanvec, sigma = Minv)
  
  # sample new sigsq
  # assumptions: IG(.5, .5) prior on sigsq
  ss = sum ( (resp - design %*% array(beta.new, dim=p) )^2 )
  n = length(resp)
  sigsq.new = rgamma(1, (n+1)/2, (ss+1)/2)
  
  
  return(list(theta.new=theta.new, 
              gamma.new=gamma.new, 
              beta.new=beta.new,
              sigsq.new=sigsq.new))
  
  
}
```

```{r mcmc over structure}

nIter = 1000

nu_1 = 5
nu_0 = 0.2

# array of parameters for each intermediate variable
# theta (1)/ gamma (p)/ beta (p)/ sigsq (1)

params = array(.1, dim = c(2, nIter, 1 + p + p + 1))


for(it in 2:nIter) {
  # sample theta
  
  
  
  ############## sample X's #####################
  new.params = sample.new(theta.old = params[1, it-1, 1],
                          gamma.old = params[1, it-1, 2:(p+1)],
                          beta.old = params[1, it-1, (p+2):(2*p+1)],
                          sigsq.old = params[1, it-1, 2*p+2],
                          design = cbind(X1, A1),
                          resp = X2)
  
  params[1, it, 1] = new.params$theta.new
  params[1, it, 2:(p+1)] = new.params$gamma.new
  params[1, it, (p+2):(2*p+1)] = new.params$beta.new
  params[1, it, 2*p+2] = new.params$sigsq.new
  
  ############## sample Z's #####################
  new.params = sample.new(theta.old = params[2, it-1, 1],
                        gamma.old = params[2, it-1, 2:(p+1)],
                        beta.old = params[2, it-1, (p+2):(2*p+1)],
                        sigsq.old = params[2, it-1, 2*p+2],
                        design = cbind(Z1, A1),
                        resp = Z2)
  
  params[2, it, 1] = new.params$theta.new
  params[2, it, 2:(p+1)] = new.params$gamma.new
  params[2, it, (p+2):(2*p+1)] = new.params$beta.new
  params[2, it, 2*p+2] = new.params$sigsq.new
  
}

```


```{r}
gams = data.frame(params[1,-1,2:3])
gams %>% group_by(X1, X2) %>% summarise(n=n()) %>% arrange(desc(n))
```


```{r}

# tweak this to take multiple predictors (or one)

# simulate data
set.seed(1)
n = 500
# binary predictor
X = cbind( rbinom(n, 1, prob = 0.4), rnorm(n) )
beta.true = c(0.2, 1)
p = length(beta.true)
y = rnorm(n, mean = X %*% beta.true)

# MCMC

nIter = 1000

nu_1 = 5
nu_0 = 0.4

theta.samples = array(0.5, dim = nIter)
gamma.samples = array(0, dim = c(nIter,p))
beta.samples = array(0, dim = c(nIter,p))
sigsq.samples = array(1, dim = nIter)

for(it in 2:nIter) {
  # sample theta
  
  new.params = sample.new(theta.samples[it-1],
                          gamma.samples[it-1,],
                          beta.samples[it-1,],
                          sigsq.samples[it-1],
                          design = X,
                          resp = y)
  
  theta.samples[it] = new.params$theta.new
  gamma.samples[it,] = new.params$gamma.new
  beta.samples[it,] = new.params$beta.new
  sigsq.samples[it] = new.params$sigsq.new
  
}


# for(it in 2:nIter) {
#   # sample theta
#   theta.samples[it] = rbeta(1, shape1 = sum(gamma.samples[it-1,]) + 1  ,shape2 = p - sum(gamma.samples[it-1,]) + 1 )
#   
#   # sample gamma
#   pg = rep(0,p) # hold the gamma probabilities
#   
#   for(pred in 1:p) {
#     
#     a = (theta.samples[it] * dnorm(beta.samples[it-1,p],mean = 0, sd = nu_1))
#     b = (1-theta.samples[it]) * dnorm(beta.samples[it-1,p],mean = 0, sd = nu_0)
#     pg[pred] = a/(a+b)
#     
#   }
#   
#   gamma.samples[it,] = rbinom(p, 1, prob = pg)
#   
#   # sample beta
#   M =  t(X) %*% X * ( sigsq.samples[it-1]^(-1) ) + diag ( ifelse(gamma.samples[it,] == 1, 1/nu_1, 1/nu_0)) 
#   beta.samples[it,] = rmvnorm(1, mean = sigsq.samples[it-1]^(-1) * solve(M) %*% t(X) %*% y, sigma = solve(M) )
#   
#   # sample sigsq
#   ss = sum ( (y - X %*% beta.samples[it,])^2 )
#   sigsq.samples[it] = 1/rgamma(1, shape = (n+1)/2, rate = (ss + 1)/2  )
#   
# }

summary(lm(y~X))

hist(beta.samples[2:nIter,2])
hist(gamma.samples[,1])
```

open questions
- multiple time points "rules"
- good choice of hyperparameters



