---
title: "sim.5.28"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## One outcome/ One predictor

\begin{align}
\sigma^2 &\sim \text{IG}(1/2,1/2) \\
P(\theta) &= {1}(0<\theta<1) \\
p(\lambda=1 \vert \theta) &= \theta \\
p(\beta \vert \lambda) &= \lambda \mathcal{ N}(0,\nu_1) + (1-\lambda)\mathcal{N}(0,\nu_0) \\
p(y \vert \beta, \mathbf{A}_0) &= \mathcal{N}(\mathbf{A_0} \beta, \sigma^2 \mathbf{I}) \\
\end{align}



One binary predictor SSVS

$$p(\theta, \lambda, \beta, \sigma^2 \vert \textbf{Data})$$

\begin{align}
p(\theta \vert \gamma) &\sim \text{Beta}(\gamma + 1, 2 - \gamma) \\
p(\gamma \vert \beta, \theta) &= \frac{\theta p(\beta \vert \gamma=1)}{\theta p(\beta \vert \gamma=1)+(1-\theta) p(\beta \vert \gamma=0)} \\
&= \frac{\theta N(0,\nu_1)}{\theta N(0,\nu_1) + (1-\theta) N(0, \nu_0)} \\
p(\beta \vert \gamma) &\sim N( (\sigma^{-2}) \mathbf{M}^{-1} X^\top y, \mathbf{M}^{-1}) \\
p(\sigma^2 \vert \cdot) &\sim \text{IG}((n+1)/2, (\text{ss}+1)/2)
\end{align}

where $$\mathbf{M} =(\sigma^{-2}X^\top X + D(\lambda)^{-1})$$

```{r}
# simulate data
set.seed(1)
n = 500
# binary predictor
A = rbinom(n, 1, prob = 0.4)
y = rnorm(n, mean = 0.5 * A)

# MCMC
# to do: figure out the regularization


nIter = 1000

nu_1 = 5
nu_0 = 0.2

gamma.samples = array(0, dim = nIter)
theta.samples = array(0.5, dim = nIter)
beta.samples = array(0, dim = nIter)
sigsq.samples = array(1, dim = nIter)

for(it in 2:nIter) {
  # sample theta
  theta.samples[it] = rbeta(1, shape1 = gamma.samples[it-1] + 1  ,shape2 = 2 - gamma.samples[it-1])
  
  # sample gamma
  pg = (theta.samples[it] * dnorm(beta.samples[it-1],mean = 0, sd = nu_1))/( (theta.samples[it] * dnorm(beta.samples[it-1],mean = 0, sd = nu_1)) + (1-theta.samples[it]) * dnorm(beta.samples[it-1],mean = 0, sd = nu_0)) 
  
  gamma.samples[it] = rbinom(1, 1, prob = pg)
  
  # sample beta
  m = (sum(A)/sigsq.samples[it-1] + ifelse(gamma.samples[it] == 1, 1/nu_1, 1/nu_0))
  beta.samples[it] = rnorm(1, mean = m^-1/sigsq.samples[it-1] * t(A) %*% y, sd = sqrt(1/m) )
  
  # sample sigsq
  ss = sum ( (y - beta.samples[it] * A)^2 )
  sigsq.samples[it] = 1/rgamma(1, shape = (n+1)/2, rate = (ss + 1)/2  )
  
}

summary(lm(y~A))

hist(beta.samples[2:nIter])
hist(gamma.samples)
```





